HMMotion: Using tracking data to predict coverage strategies in American Football

### 2024-12-30 submission.Rmd
.

## Introduction
Deciding between man and zone coverage is one of the most critical strategic choices a defensive coordinator must make before each offensive play in American football. While experienced offensive coordinators and quarterbacks often rely on visual cues to identify these defensive schemes, the increasing availability of player tracking data offers a new avenue to uncover and analyze these tactics. A notable example is Amazon's NFL Next Gen Stats model, which delivers predictions during live broadcasts (see a snapshot of the 2024 Week 12 matchup between the Pittsburgh Steelers and Cleveland Browns. These predictions, however, are often limited to analyzing specific plays without incorporating pre-snap motion which is a crucial element of modern offensive strategies. `r colorize("Ich weiß nicht, ob ich das so formulieren würde... Wir wissen nicht was deren Modell alles inkludiert. Es kann gut sein, dass sie auch tracking und daher auch premotion Daten nehmen. Oder glaubst Du ist das eine Expertise von einer Person mit domain knowledge und kein statistisches Modell?","red")`

@@

##### Feature engineering
To accurately forecast the defensive scheme (man- or zone defense) for every play, we need to create various features derived from the tracking data. In particular, we conducted the following feature engineering steps: We first consider all 11 players on each side of the field and compute features related to the convex hull of the positions of the players. In particular, for defense and offense, we compute the area spanned by the convex hull of all player such as well as the largest $y$ distance (i.e. the width of the hull) and the largest $x$ distance (i.e. the length of the hull). In addition, we select the 5 most relevant players on each side of the field (for offense: we omit the offensive line and the QB; for defense: `r colorize("TODO, Rouven","red")`) and use their standardized $x$ and $y$ coordinates as covariates. (`r colorize("Distances, orientation?? Appenix??","red")`)

@@

#### 1. Pre-motion prediction
We train different model (LASSO, Random Forest, XGBoost) to predict whether the defense plays a man- or zone coverage scheme. In particular, .....

@@

#### 2. HMM analysis
We model the movements of defensive players during the phase of pre-snap motion within a hidden Markov framework, in which the underlying states represent the offensive players to be guarded (see Franks et al. 2015 for a similar approach in basketball).

@@

The following video displays a touchdown from the Kansas City Chiefs against the Arizona Cardinals in Week 1 of the 2022 NFL season. We can see that, pre-snap, Mecole Hardman (KC #17) is in motion. He is immediately followed by the defender Marco Wilson (AZ #20), which is a clear indication for man-coverage.

@@

\textbf{den folgenden Teil könnte man auch in den Appendix packen}

A hidden Markov models consists of an observed time series $\{\boldsymbol{y}_t\}_{t=1}^T$ --- here, the y-coordinates of the defensive players and an unobserved first-order Markov chain $\{ g_t\}_{t=1}^T$, with $g_t \in \{1,\ldots,N\}$ which proxies the offensives players to be guarded at every time point $t$. The Markov chain is fully described by an initial distribution $\boldsymbol{\delta}=\bigl( \Pr(g_1=1), \ldots, \Pr(g_1=N) \bigr)$ and a transition probability matrix (t.p.m.) $\boldsymbol{\Gamma} = (\gamma_{ij}),$ with $\gamma_{ij} = \Pr(g_t = j| g_{t-1} = i), \ i,j = 1, \ldots, N$. The connection of both stochastic processes arises from the assumption that the distribution of the observations $\boldsymbol{y}_t$ are fully determined by the currently active state, i.e.\ 
\begin{equation*}
f(\boldsymbol{y}_t|g_1, \ldots, g_T, \boldsymbol{y}_1, \ldots, \boldsymbol{y}_{t-1},\boldsymbol{y}_{t+1},\ldots,\boldsymbol{y}_T) = f(\boldsymbol{y}_t|g_t).
\end{equation*}
In general, $f$ can be any density or probability mass function depending on the type of data. Following the approaches of Franks et al. (2015), we opt for a Gaussian distribution.

@@

A drawback in our approach is that we have seen that the prediction accuracy of the pre-motion model is not perfect, among other things, due to insufficient hyperparameter tuning. However, this project's focus was on the pre-snap motion, in particular the appropriate translation of this information into a statistical framework, here, the hidden Markov model. Thus, our pre-motion model can easily be replaced by another model (such as by the NFL Next Gen Stats model) which in turn can be interconnected with our modeling extension, and, thus, exploiting the information of player movements.

wir brauchen mehr Daten weil Anzahl an Plays sind nicht besonders hoch

covariates in state process and state dependendt process of the HMM

@@

#*Groom S, Morris D, Anderson L, Wang S (2024). Modeling Defensive Dynamics in Football: A Hidden Markov Model-Based Approach for Man-Marking and Zonal Defending Corner Analysis. The 2nd International Workshop on Intelligent Technologies for Precision Sports Science

@@

>>>>

## Introduction
Deciding between man and zone coverage is one of the most critical strategic choices a defensive coordinator must make before each offensive play in American football. While experienced offensive coordinators and quarterbacks often rely on visual cues to identify these defensive schemes, the increasing availability of player tracking data offers a new avenue to uncover and analyze these tactics. A notable example is Amazon's NFL Next Gen Stats model, which delivers predictions during live broadcasts (see a snapshot of the 2024 Week 12 matchup between the Pittsburgh Steelers and Cleveland Browns. These predictions, however, seem to be, from a viewership's assessment, limited to specific plays without incorporating pre-snap motion which is a crucial element of modern offensive strategies. `r colorize("Ich weiß nicht, ob ich das so formulieren würde... Wir wissen nicht was deren Modell alles inkludiert. Es kann gut sein, dass sie auch tracking und daher auch premotion Daten nehmen. Oder glaubst Du ist das eine Expertise von einer Person mit domain knowledge und kein statistisches Modell? Rouven: Ne glaube das ist ein stat. Modell aber habe mir ein paar Szenen angeschaut und da kam die Prediction eigentlich immer nur wenn sich keiner bewegte.","red")`

@@

##### Feature engineering
To accurately forecast the defensive scheme (man- or zone defense) for every play, we need to create various features derived from the tracking data. In particular, we conducted the following feature engineering steps: We first consider all 11 players on each side of the field and compute features related to the convex hull of the positions of the players. In particular, for defense and offense, we compute the area spanned by the convex hull of all player such as well as the largest $y$ distance (i.e.\ the width of the hull) and the largest $x$ distance (i.e. the length of the hull). In addition, we select the five most relevant players on each side of the field. For offense, we omit the offensive line and the QB, while, for defense, we omit nose tackles, defensive tackles and defensive ends, and select the five defenders that were the closest to the five attackers corresponding to a weighted euclidean distance, putting much more emphasis on the y-axis. Finally, we use their standardized $x$ and $y$ coordinates as covariates and order defensive and offensive players according to their $y$ coordinates.. (`r colorize("Distances, orientation?? Appenix?? Rouven: würde sagen ja","red")`)

@@

#### 1. Pre-motion prediction
We train different models (LASSO, Random Forest, XGBoost) to predict whether the defense plays a man- or zone coverage scheme. In particular, .....

@@

#### 2. Hidden Markov model
We describe the movements of the five defensive players during the phase of pre-snap motion within a hidden Markov model (see the Appendix for a thorough description). In particular, we assume the defender's $y$-coordinates in each time step $t$ to be realizations of Gaussian distributions with means according to the $y$-coordinates of the offensive players to be guarded (these are the underlying states) and a standard deviation to be estimated (see Franks et al. 2015 for a similar approach in basketball).

@@

The results of the HMM are exemplified using the following video. It displays a touchdown from the Kansas City Chiefs against the Arizona Cardinals in Week 1 of the 2022 NFL season. We can see that, pre-snap, Mecole Hardman (KC #17) is in motion. He is immediately followed by the defender Marco Wilson (AZ #20), which is a clear indication for man-coverage.

@@

Using t

@@

One limitation of our approach lies in the imperfect prediction accuracy of the pre-motion model, primarily due to insufficient hyperparameter tuning and the relatively small number of plays involving motion. However, the primary focus of this project was on pre-snap motion, particularly on how to effectively translate this information into the hidden Markov model. Importantly, our pre-motion model is modular and can be seamlessly replaced by another model, such as the NFL Next Gen Stats model, which can then be integrated into our modeling extension to fully leverage the insights provided by pre-snap motion.

@@

@@

#### Hidden Markov Model
A hidden Markov models consists of an observed time series $\{\boldsymbol{y}_t\}_{t=1}^T$ --- here, the y-coordinates of the defensive players and an unobserved first-order Markov chain $\{ g_t\}_{t=1}^T$, with $g_t \in \{1,\ldots,N\}$ which proxies the offensives players to be guarded at every time point $t$. The Markov chain is fully described by an initial distribution $\boldsymbol{\delta}=\bigl( \Pr(g_1=1), \ldots, \Pr(g_1=N) \bigr)$ and a transition probability matrix (t.p.m.) $\boldsymbol{\Gamma} = (\gamma_{ij}),$ with $\gamma_{ij} = \Pr(g_t = j| g_{t-1} = i), \ i,j = 1, \ldots, N$. The connection of both stochastic processes arises from the assumption that the distribution of the observations $\boldsymbol{y}_t$ are fully determined by the currently active state, i.e.\ 
\begin{equation*}
f(\boldsymbol{y}_t|g_1, \ldots, g_T, \boldsymbol{y}_1, \ldots, \boldsymbol{y}_{t-1},\boldsymbol{y}_{t+1},\ldots,\boldsymbol{y}_T) = f(\boldsymbol{y}_t|g_t).
\end{equation*}
In general, $f$ can be any density or probability mass function depending on the type of data. Following the approaches of Franks et al. (2015), we opt for a Gaussian distribution.
